---
title: "STAT 118: Introductory Statistics"
subtitle: "Inference for Numerical Data (One Group)"
author: "Anthony Scotina"
date: 
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: my-theme.css
    nature:
      countIncrementalSlides: false
      highlightLines: true
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_accent(base_color = "#bf67bc") #3E8A83?
```

```{r, include = FALSE}
library(tidyverse)
library(mosaic)
library(moderndive)
library(nycflights13)
library(openintro)
set.seed(12)
data(COL)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      echo = TRUE, dpi = 300)

airbnb_boston <- read_csv("~/Dropbox/Teaching/03-Simmons Courses/STAT118-Introductory Statistics [ONLINE]/Data/airbnb_boston.csv")
airbnb_boston = airbnb_boston %>%
  filter(room_type != "Hotel room")

theme_set(theme_bw() +
  theme(axis.title.x = element_text(size = 14, face = "bold"), 
        axis.title.y = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12, face = "bold"), 
        axis.text.y = element_text(size = 12, face = "bold")))
```

<!--
pagedown::chrome_print("~/Dropbox/Teaching/03-Simmons Courses/MATH228-Introduction to Data Science/Lecture Slides/01-Introduction/01-Introduction.html")
-->

class: center, middle, frame

# Module Plan

## Central Limit Theorem .display[for means]
    
--

## The .display[t-distribution]

--

## Hypothesis testing for .display[single mean]
    
---

class: center, middle, frame

# Central Limit Theorem (for means)

---

# Boston Airbnb Data

```{r}
airbnb_boston
```

- Sample of $n=2980$ Boston Airbnbs

We are interested in estimating the **average price per night** of Airbnbs in *major US cities*.

---

# Exploratory Data Analysis

.pull-left[
```{r}
gf_histogram( ~ price, 
              data = airbnb_boston)
```

- **right-skewed**
- **typical** prices per night are ~150

]

--

.pull-right[
```{r, comment = ""}
favstats( ~ price, 
          data = airbnb_boston)
```

- $\bar{x}=163.52$ (**sample mean**)

- $s=140.83$ (**sample standard deviation**)
]

---

# Sampling Variability

In our *single sample* of Airbnbs, 

- $\bar{x}=163.52$, $s=140.83$

--

**Central Limit Theorem**: When we collect a *sufficiently large* sample of $n$ *independent* observations, the **sampling distribution** of $\bar{x}$ is Normal with:
- $\mu_{\bar{x}}=\mu$ (*unknown* population mean)
- $\sigma_{\bar{x}}=\sigma/\sqrt{n}$

--

.center[
```{r, echo = FALSE, fig.height = 4, out.width = "65%"}
set.seed(12)
samp = data.frame(var = rnorm(10000))
gf_density( ~ var, data = samp) + 
  geom_histogram(binwidth = 0.2, alpha = 0.3, fill = "blue") + 
  labs(x = expression(bar(x)), y = "",
       title = "Sampling Distribution", subtitle = "(Sample Mean)") + 
  scale_x_continuous(labels = c("", "", expression(mu), "", "")) +
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())
```
]

---

# CLT Conditions

Two conditions for using the CLT with $\bar{x}$

**Independence**: The sampled observations are independent. 

- Satisfied if the sample is **random**

    - Is this condition satisfied for the Airbnb data?

--

**Normality**: *IF the sample is small*, the sample observations must come from a *Normal population*. 

- $n<30$: If there are *no clear outliers*, we **assume** the data are sampled from a *nearly Normal* distribution. 

- $n\geq30$: If there are no **extreme outliers**, we assume that the sampling distribution of $\bar{x}$ is nearly Normal. 
    
---

# Standard Error

The **standard error** describes how much estimated variability one can expect in the *sample statistic*. 

.center[
```{r, echo = FALSE, fig.width = 5, fig.asp = .32, out.width = "60%"}
data.frame(range = seq(0, 200, 1)) %>%
  ggplot(aes(x = range)) + 
  geom_blank() + 
  annotate("point", x = 163.52, y = 0.25, size = 3) + 
  geom_segment(y = 0.25, yend = 0.25, 
               x = 158.4617, xend = 168.5783, 
               color = "red", lty = 2) + 
  scale_x_continuous(labels = scales::dollar, limits = c(150, 175)) +
  theme_minimal() + 
  labs(x = "", y = "", title = "95% Confidence Interval") + 
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.text.x = element_text(size = 12, face = "bold"))
```
]

--

The **standard deviation** of $\bar{x}$ is given by $SD(\bar{x})=\sigma/\sqrt{n}$. 

- We *estimate* this with the **standard error**: $$SE_{\bar{x}}=\frac{s}{\sqrt{n}}=\frac{140.83}{\sqrt{2980}}=2.58$$

--

```{r}
140.83/sqrt(2980)
```

---

class: center, middle, frame

# The t-distribution

---

# The t-distribution

.center[
```{r, echo = FALSE, fig.asp = .52, out.width = "65%"}
plot(c(-5, 5),
     c(0, dnorm(0)),
     type = 'n',
     axes = FALSE, xlab = "", ylab = "")
axis(1, seq(-6, 6, 2))
abline(h = 0)

xleg <- 2
yleg <- 0.35
yleg.line.offset <- -0.07
line.leg.width <- 0.55

lines(
    c(xleg, xleg + line.leg.width),
    rep(yleg + yleg.line.offset, 2),
    col = COL[1], lty = 1, lwd = 1.8)
text(xleg + line.leg.width, yleg + yleg.line.offset,
    "t-distribution",
    col = COL[1], pos = 4)

X <- seq(-6, 6, 0.01)

Y <- dt(X, 2)
lines(X, Y, lwd = 1.8, col = COL[1])

```
]

--

.center[
```{r, echo = FALSE, fig.asp = .52, out.width = "65%"}
plot(c(-5, 5),
     c(0, dnorm(0)),
     type = 'n',
     axes = FALSE, xlab = "", ylab = "")
axis(1, seq(-6, 6, 2))
abline(h = 0)

xleg <- 2
yleg <- 0.35
yleg.line.offset <- -0.07
line.leg.width <- 0.55
lines(
    c(xleg, xleg + line.leg.width),
    rep(yleg, 2),
    col = COL[4], lty = 3, lwd = 2.5)
lines(
    c(xleg, xleg + line.leg.width),
    rep(yleg + yleg.line.offset, 2),
    col = COL[1], lty = 1, lwd = 1.8)
text(xleg + line.leg.width, yleg,
    "Standard Normal",
    col = COL[4], pos = 4)
text(xleg + line.leg.width, yleg + yleg.line.offset,
    "t-distribution",
    col = COL[1], pos = 4)

X <- seq(-6, 6, 0.01)
Y <- dnorm(X)
lines(X, Y, lty = 3, lwd = 2.5, col = COL[4])

Y <- dt(X, 2)
lines(X, Y, lwd = 1.8, col = COL[1])

```
]

---

# The t-distribution

.center[
```{r, echo = FALSE, fig.asp = .52, out.width = "65%"}
plot(c(-5, 5),
     c(0, dnorm(0)),
     type = 'n',
     axes = FALSE, xlab = "", ylab = "")
axis(1, seq(-6, 6, 2))
abline(h = 0)

xleg <- 2
yleg <- 0.35
yleg.line.offset <- -0.07
line.leg.width <- 0.55
lines(
    c(xleg, xleg + line.leg.width),
    rep(yleg, 2),
    col = COL[4], lty = 3, lwd = 2.5)
lines(
    c(xleg, xleg + line.leg.width),
    rep(yleg + yleg.line.offset, 2),
    col = COL[1], lty = 1, lwd = 1.8)
text(xleg + line.leg.width, yleg,
    "Standard Normal",
    col = COL[4], pos = 4)
text(xleg + line.leg.width, yleg + yleg.line.offset,
    "t-distribution",
    col = COL[1], pos = 4)

X <- seq(-6, 6, 0.01)
Y <- dnorm(X)
lines(X, Y, lty = 3, lwd = 2.5, col = COL[4])

Y <- dt(X, 2)
lines(X, Y, lwd = 1.8, col = COL[1])

```
]

- **Always** centered at *zero*

- One parameter controls the *form* of the bell-shape: the **degrees of freedom** (*df*)
    - *df* $=n-1$
    
---

# Degrees of Freedom (df)

*df* $=n-1$

- The *larger the sample*, the **larger the df**. 

--

.center[
```{r, echo = FALSE, fig.asp = .52, out.width = "75%"}
plot(c(-5, 5),
     c(0, dnorm(0)),
     type = 'n',
     axes = FALSE, xlab = "", ylab = "")
at <- seq(-10, 10, 2)
axis(1, at)
axis(1, at - 1, rep("", length(at)), tcl = -0.1)
abline(h = 0)

COL. <- fadeColor(COL[1], c("FF", "89", "68", "4C", "33"))
COLt <- fadeColor(COL[1], c("FF", "AA", "85", "60", "45"))
DF   <- c('Normal', 8, 4, 2, 1)

X <- seq(-10, 10, 0.02)
Y <- dnorm(X)
lines(X, Y, col = COL.[1])

for (i in 2:5) {
  Y <- dt(X, as.numeric(DF[i]))
  lines(X, Y, col = COL.[i], lwd = 1.5)
}

legend(2.5, 0.4,
       legend = c(DF[1],
       paste('t, df = ', DF[2:5], sep = '')),
       col = COL.,
       text.col = COLt,
       lty = rep(1, 5),
       lwd = 1.5)
```
]

--

In the Airbnb example, $n=2980$, so *df* $=2979$. 

---

# Confidence Intervals using t-distribution

Define $\mu$ as the **population mean** of Airbnb price per night. 

- This is *unknown*, but $\bar{x}=163.52$ serves as a **point estimate**. 

--

A **confidence interval** represents a *range of plausible values* for $\mu$. 

Using the **t-distribution**:

- $\bar{x}\pm t^{*}_{df}\times SE_{\bar{x}}$

--

The $t^{*}_{df}$ value depends on the **confidence level** AND the *df*. 

.center[
```{r, echo = FALSE, fig.height = 3, out.width = "75%"}
par(mar = c(3.3, 1, 0.15, 1), mgp = c(2.1, 0.6, 0))
normTail(M = c(-1.96, 1.96),
         df = 2979,
         col = COL[1],
         xlim = 3.3 * c(-1, 1),
         ylim = c(0, 0.59),
         xlab="t-distribution, df = 2979")
X <- rev(seq(-4, 4, 0.025))
Y <- dt(X, 10) # makes better visual

yMax <- 0.41

lines(1.96*c(-1,1), rep(yMax,2), lwd=2)
lines(rep(-1.96,2), c(0,yMax), lty=2, col=COL[6])
lines(rep( 1.96,2), c(0,yMax), lty=2, col=COL[6])
text(0, yMax, '95%, extends -1.96 to 1.96', pos=3)
```
]

---

# Confidence Intervals using t-distribution

Define $\mu$ as the **population mean** of Airbnb price per night. 

- This is *unknown*, but $\bar{x}=163.52$ serves as a **point estimate**. 

A **confidence interval** represents a *range of plausible values* for $\mu$. 

Using the **t-distribution**:

- $\bar{x}\pm t^{*}_{df}\times SE_{\bar{x}}$

The $t^{*}_{df}$ value depends on the **confidence level** AND the *df*. 

.center[
```{r, echo = FALSE, fig.height = 3, out.width = "75%"}
par(mar = c(3.3, 1, 0.15, 1), mgp = c(2.1, 0.6, 0))
normTail(M = c(-2.144787, 2.144787),
         df = 14,
         col = COL[1],
         xlim = 3.3 * c(-1, 1),
         ylim = c(0, 0.59),
         xlab="t-distribution, df = 14")
X <- rev(seq(-4, 4, 0.025))
Y <- dt(X, 10) # makes better visual

yMax <- 0.41

lines(2.144787*c(-1,1), rep(yMax,2), lwd=2)
lines(rep(-2.144787,2), c(0,yMax), lty=2, col=COL[6])
lines(rep( 2.144787,2), c(0,yMax), lty=2, col=COL[6])
text(0, yMax, '95%, extends -2.145 to 2.145', pos=3)
```
]

---

# qt()

We'll use `qt()` to find $t_{df}$ distribution *quantiles*.
- Works similarly to `qnorm()`. 

.pull-left[
```{r}
val = 1 - (1-0.95)/2 # For 95% CI
qt(val, df = 2979) 
```

```{r}
val = 1 - (1-0.95)/2 # For 95% CI
qt(val, df = 14)
```
]

--

.pull-right[
```{r}
qnorm(val)
```

```{r}
qt(val, 
   df = c(5, 50, 250, 250000))
```
]

---

# 95% Confidence Intervals

**Airbnb example**

- $SE_{\bar{x}}=s/\sqrt{n}=140.83/\sqrt{2980}=2.58$

- $\bar{x}\pm t^{*}_{df}\times SE_{\bar{x}} \longrightarrow 163.52\pm1.96\times2.58$

```{r}
c(163.52 - 1.96*2.58, 163.52 + 1.96*2.58)
```

--

**Interpretation**: We are **95% confident** that the *average* price per night of Airbnbs in *major US cities* is between 	&#36;158.46 and &#36;168.58. 

.center[
```{r, echo = FALSE, fig.width = 5, fig.asp = .32, out.width = "60%"}
data.frame(range = seq(0, 200, 1)) %>%
  ggplot(aes(x = range)) + 
  geom_blank() + 
  annotate("point", x = 163.52, y = 0.25, size = 3) + 
  geom_segment(y = 0.25, yend = 0.25, 
               x = 158.4617, xend = 168.5783, 
               color = "red", lty = 2) + 
  scale_x_continuous(labels = scales::dollar, limits = c(150, 175)) +
  theme_minimal() + 
  labs(x = "", y = "", title = "95% Confidence Interval") + 
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.text.x = element_text(size = 12, face = "bold"))
```
]

---

class: center, middle, frame

# Hypothesis Testing (for single mean)

---

# Do you get enough sleep

[According to the NIH](https://www.nhlbi.nih.gov/files/docs/public/sleep/healthysleepfs.pdf), most adults require *~7.5 hours of sleep* each night. 

--

```{r, echo = FALSE}
set.seed(12)
sleep_samp = data.frame(hours = rnorm(25, mean = 6.3, sd = 2.3))
```

In a survey of 25 randomly selected college students...

```{r}
favstats( ~ hours, data = sleep_samp)
```

- $\bar{x}=5.866766$ hours, $s=2.193591$ hours

--

<br>

**Question**: Do the data provide sufficient *evidence* that college students generally sleep for *fewer* hours than recommended?

---

# Exploratory Data Analysis

```{r, out.width = "45%"}
gf_histogram(~ hours, binwidth = 1, data = sleep_samp)
```

- **unimodal**
- *no major outliers*

---

# Competing Hypotheses

Define $\mu$ as the **population mean** hours of sleep per night. 

- $H_{0}$ (**Null hypothesis**): $\mu = 7.5$

- $H_{A}$ (**Alternative hypothesis**): $\mu < 7.5$

--

We'll play the role of the **skeptic**. Assume $H_{0}$ until there is *convincing evidence* that it is false. 

- Calculate the **test statistic**, and see how that "evidence" appears *in a world where the null is true*. 

---

# Test Statistic

Similar to hypothesis tests for **proportions**!

- **Proportions**: $$Z=\frac{\text{point estimate} - \text{null value}}{SE_{\hat{p}}}$$

--

- **Means**: $$T=\frac{\text{point estimate} - \text{null value}}{SE_{\bar{x}}}$$
    - $SE_{\bar{x}}=s/\sqrt{n}$
    
--

```{r}
T = (5.866766 - 7.5)/(2.193591/sqrt(25))
T
```

---

# Null Distribution

.center[
```{r, echo = FALSE, fig.height = 4, out.width = "65%"}
samp = data.frame(var = rt(10000, df = 24))
gf_density( ~ var, data = samp) + 
  geom_histogram(binwidth = 0.2, alpha = 0.3, fill = "blue") + 
  geom_vline(xintercept = -3.64, color = "red") + 
  labs(x = "T",
       y = "",
       title = "Null Distribution", 
       subtitle = "t-distribution with 24 degrees of freedom") + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())
```
]

- **p-value**: The probability of observing data *more extreme than yours*, **in a world where the null hypothesis is true**

--

Calculate **tail areas** in the *t*-distribution with `pt()`:
```{r}
pt(-3.72274, df = 24)
```

---

# t.test

These *hypothesis tests for a single mean* commonly go by a different name:
- **One-sample t-test**

--

**In R**:

```{r}
t.test( ~ hours, alternative = "less", mu = 7.5, 
        data = sleep_samp)
```

--

**Conclusion**: p-value $<0.05$, *reject the null*. We have evidence to suggest that students sleep for fewer hours per night than the recommended amount, on average. 

