---
title: "STAT 118: Introductory Statistics"
subtitle: "Inference for Numerical Data (Two Groups)"
author: "Anthony Scotina"
date: 
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: my-theme.css
    nature:
      countIncrementalSlides: false
      highlightLines: true
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_accent(base_color = "#bf67bc") #3E8A83?
```

```{r, include = FALSE}
library(tidyverse)
library(mosaic)
library(moderndive)
library(nycflights13)
library(openintro)
set.seed(12)
data(COL)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      echo = TRUE, dpi = 300)

airbnb_boston <- read_csv("~/Dropbox/Teaching/03-Simmons Courses/STAT118-Introductory Statistics [ONLINE]/Data/airbnb_boston.csv")
airbnb_boston = airbnb_boston %>%
  filter(room_type != "Hotel room")

theme_set(theme_minimal() +
  theme(axis.title.x = element_text(size = 14, face = "bold"), 
        axis.title.y = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12, face = "bold"), 
        axis.text.y = element_text(size = 12, face = "bold")))
```

<!--
pagedown::chrome_print("~/Dropbox/Teaching/03-Simmons Courses/MATH228-Introduction to Data Science/Lecture Slides/01-Introduction/01-Introduction.html")
-->

class: center, middle, frame

# Module Plan

## Confidence intervals for .display[two means]
    
--

## Hypothesis testing for .display[two means]
    
---

class: center, middle, frame

# Comparing Groups (again)

---

# Jam Session

[Spotify data](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-21/readme.md)

- A *sample* of 10,458 songs from Spotify

- Songs categorized as either **rock** or **pop**

```{r, echo = FALSE}
spotify_songs <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')
spotify_songs %>%
  select(track_artist, playlist_genre, danceability) %>%
  filter(playlist_genre %in% c("rock", "pop")) 
```

--

**Question**: Is there a *relationship* between `genre` and `danceability`?

---

# Exploratory Data Analysis

**Summary statistics**

```{r, echo = FALSE}
spotify_songs %>%
  select(track_artist, playlist_genre, danceability) %>%
  filter(playlist_genre %in% c("rock", "pop")) %>%
  group_by(playlist_genre) %>%
  summarize(mean_dance = mean(danceability), 
            sd_dance = sd(danceability), 
            n = n())
```

.pull-left[
**Data visualization**

```{r, echo = FALSE, fig.width = 8, fig.height = 5}
spotify_rp = spotify_songs %>%
  select(track_name, track_artist, playlist_genre, danceability) %>%
  filter(playlist_genre %in% c("rock", "pop"))

gf_boxplot(danceability ~ playlist_genre, data = spotify_rp) + 
  labs(x = "")
```
]

.pull-right[
It *appears* that **pop** songs tend to have *higher* measures of "danceability" compared to **rock** songs. 

- Is the difference **large enough**?
]

---

# Point Estimate

**Difference between means**

- $\bar{x}_{p}-\bar{x}_{r}=0.639-0.521=0.118$

--

**Interpretation**: *In the sample*, the *danceability score* of pop songs is 0.118 points higher *on average* than rock songs. 

- But **samples vary**!

--

.pull-left[
```{r, echo = FALSE, fig.width = 5, fig.asp = .32}
data.frame(range = seq(0, 1, 0.01)) %>%
  ggplot(aes(x = range)) + 
  geom_blank() + 
  annotate("point", x = 0.118, y = 0.25, size = 3) + 
  scale_x_continuous(limits = c(0.1, 0.15)) +
  theme_minimal() + 
  labs(x = "", y = "", title = "Point Estimate") + 
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.text.x = element_text(size = 12, face = "bold"))
```
]

.pull-right[
```{r, echo = FALSE, fig.width = 5, fig.asp = .32}
data.frame(range = seq(0, 1, 0.01)) %>%
  ggplot(aes(x = range)) + 
  geom_blank() + 
  annotate("point", x = 0.1187537, y = 0.25, size = 3) + 
  geom_segment(y = 0.25, yend = 0.25, 
               x = 0.1135911, xend = 0.1239164, 
               color = "red", lty = 2) + 
  scale_x_continuous(limits = c(0.1, 0.15)) +
  theme_minimal() + 
  labs(x = "", y = "", title = "Point Estimate (plus 95% Confidence Interval)") + 
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.text.x = element_text(size = 12, face = "bold"))
```
]

---

# The t-distribution

Similar to the *one sample* case for **means**, we'll model the *difference between means* using the **t-distribution**. 

.center[
```{r, echo = FALSE, fig.asp = .52, out.width = "65%"}
plot(c(-5, 5),
     c(0, dnorm(0)),
     type = 'n',
     axes = FALSE, xlab = "", ylab = "")
axis(1, seq(-6, 6, 2))
abline(h = 0)

xleg <- 2
yleg <- 0.35
yleg.line.offset <- -0.07
line.leg.width <- 0.55
lines(
    c(xleg, xleg + line.leg.width),
    rep(yleg, 2),
    col = COL[4], lty = 3, lwd = 2.5)
lines(
    c(xleg, xleg + line.leg.width),
    rep(yleg + yleg.line.offset, 2),
    col = COL[1], lty = 1, lwd = 1.8)
text(xleg + line.leg.width, yleg,
    "Standard Normal",
    col = COL[4], pos = 4)
text(xleg + line.leg.width, yleg + yleg.line.offset,
    "t-distribution",
    col = COL[1], pos = 4)

X <- seq(-6, 6, 0.01)
Y <- dnorm(X)
lines(X, Y, lty = 3, lwd = 2.5, col = COL[4])

Y <- dt(X, 2)
lines(X, Y, lwd = 1.8, col = COL[1])
```
]

--

**Conditions**

- *Independence, extended*

- *Normality*

---

# Reviewing Standard Error

The **standard error** gives a measure of *variability* for sample statistics from *sample to sample*. 

- For *single mean*: $$SE_{\bar{x}}=\frac{s}{\sqrt{n}}=\sqrt{\frac{s^{2}}{n}}$$

--

- For *difference between means*: $$SE_{\bar{x}_{1}-\bar{x}_{2}}=\sqrt{\frac{s_{1}^{2}}{n_{1}}+\frac{s_{2}^{2}}{n_{2}}}$$

--

**Spotify data**

```{r}
sqrt(0.128^2/5507 + 0.140^2/4951)
```

---

# 95% Confidence Intervals

Similar to confidence intervals for a *single mean*: $$\bar{x}\pm t^{*}_{df}\times\frac{s}{\sqrt{n}}$$

--

For a **difference between means**: $$(\bar{x}_{1}-\bar{x}_{2})\pm t^{*}_{df}\times \sqrt{\frac{s_{1}^{2}}{n_{1}}+\frac{s_{2}^{2}}{n_{2}}}$$

--

- In this case, $df=\min(n_{1}-1, n_{2}-1)$

```{r}
min(5507-1, 4951-1)
```

---

# 95% Confidence Intervals

As before, we'll use `qt()` to find $t^{*}_{df}$. 
- Depends on the **df** *and* **confidence level**. 

```{r}
val = 1 - (1-0.95)/2 #For 95% CI
qt(val, df = 4950)
```

--

Putting it all together...

```{r}
c(0.118 - 1.960443*0.002633234, 
  0.118 + 1.960443*0.002633234)
```

--

**Interpretation**: We are **95% confident** that the average danceability of *pop* songs is between 0.113 and 0.123 points *higher* than *rock* songs. 

---

# Smaller Samples

What if we only had 10 pop and 10 rock songs?

--

.pull-left[
```{r, echo = FALSE}
set.seed(228)
spotify_rp_small = spotify_rp %>%
  group_by(playlist_genre) %>%
  sample_n(10) %>%
  ungroup()

spotify_rp_small %>%
  group_by(playlist_genre) %>%
  summarize(mean_dance = mean(danceability), 
            sd_dance = sd(danceability))
```

$\bar{x}_{p}-\bar{x}_{r}=0.092$
]

.pull-right[
```{r, echo = FALSE}
gf_boxplot(danceability ~ playlist_genre, data = spotify_rp_small) + 
  labs(x = "")
```
]

--

**95% Confidence Interval**

```{r, echo = FALSE}
c((0.667-0.575) - qt(0.975, df = 9)*sqrt(0.0976^2/10 + 0.144^2/10), 
  (0.667-0.575) + qt(0.975, df = 9)*sqrt(0.0976^2/10 + 0.144^2/10))
```

---

class: center, middle, frame

# Hypothesis Testing (for two means)

---

# (Smaller) Spotify Data

Let's work with the **smaller** Spotify dataset! 

```{r}
spotify_rp_small
```

---

# Hypotheses

Let $\mu_{p}$ and $\mu_{r}$ be the (unknown) **population mean** danceability of the two genres. 

- $H_{0}$ (**Null** hypothesis): $\mu_{p}-\mu_{r}=0$, or $\mu_{p}=p_{r}$

- $H_{A}$ (**Alternative** hypothesis): $\mu_{p}-\mu_{r}\neq0$ or $\mu_{p}\neq \mu_{r}$

--

*In words...*

- $H_{0}$: **No difference** in mean danceability between genres
    - *danceability* and *genre* are **independent**
    
- $H_{A}$: **A difference** in mean danceability between genres
    - *danceability* and *genre* are **associated**
 
<br>

We'll play the role of the **skeptic**. Assume $H_{0}$ until there is *convincing evidence* that it is false. 

- "Evidence": $\bar{x}_{p}-\bar{x}_{r}=0.092$

---

# Simulating the Null Distribution

(One shuffle)

```{r, echo = FALSE}
spotify_rp_shuffle = spotify_rp_small
spotify_rp_shuffle$playlist_genre = shuffle(spotify_rp_shuffle$playlist_genre)
```

```{r}
spotify_rp_shuffle
```

---

# Simulating the Null Distribution

(Several shuffles)

```{r}
diffmean(danceability ~ shuffle(playlist_genre), data = spotify_rp_small)
```

```{r}
diffmean(danceability ~ shuffle(playlist_genre), data = spotify_rp_small)
```

```{r}
diffmean(danceability ~ shuffle(playlist_genre), data = spotify_rp_small)
```

```{r}
diffmean(danceability ~ shuffle(playlist_genre), data = spotify_rp_small)
```

---

# Null Distribution

.pull-left[
```{r, echo = FALSE}
set.seed(118)
gen_null <- do(1000) * diffmean(danceability ~ shuffle(playlist_genre), data = spotify_rp_small)
#sum(abs(gen_null$diffmean) >= 0.092)/1000
gf_histogram(~ diffmean, data = gen_null) + 
  labs(x = "Difference in Mean danceability", 
       title = "Rock vs Pop", 
       subtitle = "Null Distribution: 1,000 simulations") + 
  geom_vline(xintercept = 0.092, color = "red") + 
  geom_vline(xintercept = -0.092, color = "red")
```
]

.pull-right[
Why **two** red lines?

- $H_{A}$: $\mu_{p}-\mu_{r}\neq0$ 

"Evidence" *in favor of* $H_{A}$ comes from *both sides*!
]

--

- In 1,000 simulations of $H_{0}$, 10.8% (108/1000) yielded a *difference in means* **more extreme** than observed. 

---

# Test Statistic

Alternatively, provided the **conditions** are met, we can model the *null distribution* using the **t-distribution**. 

- First, calculate the **test statistic**: $$T=\frac{\text{point estimate} - \text{null value}}{SE_{\bar{x}_{p}-\bar{x}_{r}}}=\frac{(\bar{x}_{p}-\bar{x}_{r})-0}{\sqrt{\frac{s_{p}^{2}}{n_{p}} + \frac{s_{r}^{2}}{n_{r}}}}$$

--

```{r}
(0.092 - 0)/sqrt(0.0976^2/10 + 0.144^2/10)
```

---

# (Standardized) Null Distribution

We can compare the T statistic, $T=1.67$ to the (standardized) null distribution. 
- **t-distribution** with $df = min(n_{p}-1, n_{r}-1)=9$ 

--

.pull-left[
```{r, echo = FALSE}
set.seed(12)
samp = data.frame(var = rt(10000, df = 9))
gf_density( ~ var, data = samp) + 
  geom_histogram(binwidth = 0.2, alpha = 0.3, fill = "blue") + 
  geom_vline(xintercept = 1.67, color = "red") + 
  geom_vline(xintercept = -1.67, color = "red") + 
  labs(x = "T",
       y = "",
       title = "Standardized Null Distribution", 
       subtitle = "t-distribution with 9 degrees of freedom") + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())
```
]

.pull-right[
```{r, echo = FALSE}
#sum(abs(gen_null$diffmean) >= 0.092)/1000
gf_histogram(~ diffmean, data = gen_null) + 
  labs(x = "Difference in Mean danceability", 
       title = "Rock vs Pop", 
       subtitle = "Null Distribution: 1,000 simulations") + 
  geom_vline(xintercept = 0.092, color = "red") + 
  geom_vline(xintercept = -0.092, color = "red")
```
]

---

# t.test

These *hypothesis tests for a difference between means* commonly go by a different name:
- **Two-sample t-test**

--

**In R**:

```{r}
t.test(danceability ~ playlist_genre, alternative = "two.sided", 
        data = spotify_rp_small)
```

--

**Conclusion**: p-value $>0.05$, *fail to reject the null*. We do not have enough evidence to suggest that average danceability score is *different* between rock and pop songs.

---

# BUT

What happens when we work with a **larger** dataset?!

```{r}
t.test(danceability ~ playlist_genre, alternative = "two.sided", 
        data = spotify_rp)
```

--

**Sample size** plays an important role. 

- Always include **confidence intervals** for the complete picture. 
